---
title: "Getting Started with DeOPUS"
author: "Ha Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DeOPUS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

DeOPUS (Deconvolution via Optimized Power-transformed Unmixing with Shrinkage)
is a reference-based cellular deconvolution method that estimates cell-type
proportions from bulk RNA sequencing data.

## Installation

Install DeOPUS from GitHub:

```{r install, eval=FALSE}
# Install devtools if not already installed
install.packages("devtools")

# Install DeOPUS
devtools::install_github("tinnlab/DeOPUS")
```

Load the package:

```{r load}
library(DeOPUS)
```

## Input Data Format

DeOPUS requires two inputs:

1. **Bulk matrix** (`bulk`): A genes × samples matrix of bulk RNA-seq expression
   values in linear scale (non-negative, not log-transformed).
2. **Reference matrix** (`reference`): A genes × cell types matrix of average
   expression profiles per cell type, typically derived from scRNA-seq.

Both matrices must have matching gene identifiers as row names.

## Quick Example

The package includes a built-in simulated benchmark dataset (`sampleData`) with
known ground-truth proportions for 2 cell types across 512 samples.

```{r sample-data, eval=FALSE}
# Load the built-in sample dataset
data(sampleData)

# Inspect dimensions
dim(sampleData$bulk)          # 11852 genes x 512 samples
dim(sampleData$cellTypeExpr)  # 11852 genes x 2 cell types

# Run deconvolution on a small subset (10 samples for speed)
set.seed(42)
idx <- sample(ncol(sampleData$bulk), 10)

results <- deconvolve(
  bulk      = sampleData$bulk[, idx],
  reference = sampleData$cellTypeExpr,
  alpha     = 0.01,
  n_cores   = 1,
  verbose   = TRUE
)

# View estimated proportions (samples x cell types)
head(results$proportions)
```

### Evaluate Against Ground Truth

`bulkRatio` row names and `cellTypeExpr` column names may be in different
orders, so align by cell-type name before computing the correlation:

```{r evaluate, eval=FALSE}
# Align bulkRatio cell-type ORDER to match results$proportions columns
ct <- colnames(results$proportions)
true_props <- t(sampleData$bulkRatio[ct, idx, drop = FALSE])  # 10 x 2

# Pearson correlation per sample
cor_values <- sapply(seq_len(nrow(results$proportions)), function(i) {
  cor(results$proportions[i, ], true_props[i, ], method = "pearson")
})

cat("Mean Pearson correlation:", round(mean(cor_values, na.rm = TRUE), 3), "\n")
```

## Method Details

DeOPUS applies a multi-level adaptive transformation:

### 1. Hierarchical Shrinkage

- **Local shrinkage** ($\tau$): Attenuates the influence of individual
  high-variance genes.
- **Global shrinkage** ($\phi$): Provides overall regularization across the
  expression profile.

### 2. Power Transformation

A variance-stabilizing power transformation handles the high dynamic range:

$$z = \frac{(q/s + \kappa)^{\mu \omega g} - \kappa^\mu}{\mu g}$$

where $q$ is the expression value, $s$ is a scale factor, $g$ is the global
shrinkage weight, and $\kappa$, $\mu$, $\omega$ are shape parameters.

### 3. Quantile Normalization

Spline-based quantile normalization ensures robust comparison between profiles.

### 4. Optimization

DeOPUS uses L-BFGS-B optimization with box constraints (proportions ≥ 0) to
minimize the loss between transformed bulk and reconstructed expression.

## Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `alpha`   | 0.01    | Regularization parameter |
| `power`   | 2       | Loss function power (1 = MAE, 2 = MSE) |
| `n_cores` | 1       | Parallel processing cores |
| `maxit`   | 100     | Maximum optimization iterations |
| `verbose` | FALSE   | Print progress messages |

## Benchmarking

The package ships benchmark and analysis scripts in `inst/scripts/`. After
installation, source them via `system.file()`:

```{r benchmark, eval=FALSE}
# Run benchmark on real datasets
source(system.file("scripts/benchmark/run_benchmark_real.R", package = "DeOPUS"))

# Generate visualization dashboard
source(system.file("scripts/analysis/visualize_results.R", package = "DeOPUS"))
```

## Session Info

```{r sessioninfo}
sessionInfo()
```
