---
title: "Getting Started with DeOPUS"
author: "Ha Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DeOPUS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

DeOPUS (Deconvolution via Optimized Power-transformed Unmixing with Shrinkage)
is a reference-based cellular deconvolution method that estimates cell-type
proportions from bulk RNA sequencing data.

## Installation

Install DeOPUS from GitHub:

```{r install, eval=FALSE}
# Install devtools if not already installed
install.packages("devtools")

# Install DeOPUS
devtools::install_github("tinnlab/DeOPUS")
```

Load the package:

```{r load}
library(DeOPUS)
```

## Input Data Format

DeOPUS requires two inputs:

1. **Bulk matrix** (`bulk`): A genes × samples matrix of bulk RNA-seq expression
   values in linear scale (non-negative, not log-transformed).
2. **Reference matrix** (`reference`): A genes × cell types matrix of average
   expression profiles per cell type, typically derived from scRNA-seq.

Both matrices must have matching gene identifiers as row names.

## Quick Example: Simulated Dataset

The package includes a built-in simulated benchmark dataset with known
ground-truth proportions for 2 cell types across 512 samples.

```{r simulated-data, eval=FALSE}
# Load built-in simulated dataset
data(simulated)

# Inspect dimensions
dim(simulated$bulk)          # 11852 genes x 512 samples
dim(simulated$cellTypeExpr)  # 11852 genes x 2 cell types

# Run deconvolution on a small subset (10 samples for speed)
set.seed(42)
idx <- sample(ncol(simulated$bulk), 10)

results <- deconvolve(
  bulk      = simulated$bulk[, idx],
  reference = simulated$cellTypeExpr,
  alpha     = 0.01,
  n_cores   = 1,
  verbose   = TRUE
)

# View estimated proportions (samples x cell types)
head(results$proportions)
```

### Evaluate Against Ground Truth

```{r evaluate-simulated, eval=FALSE}
# Ground-truth proportions for the selected samples
true_props <- t(simulated$bulkRatio[, idx])  # samples x cell types

# Pearson correlation per sample between estimated and true proportions
cor_values <- sapply(seq_len(nrow(results$proportions)), function(i) {
  cor(results$proportions[i, ], true_props[i, ], method = "pearson")
})

cat("Mean Pearson correlation:", round(mean(cor_values, na.rm = TRUE), 3), "\n")
```

## Real Dataset: GSE77343

The package also bundles a real bulk RNA-seq dataset from GEO accession
GSE77343 with 17,275 genes, 197 samples, and a reference panel of 15 cell
types. Because of its size, it is stored in `inst/extdata/` and must be
loaded explicitly.

```{r real-data, eval=FALSE}
# Load real GEO dataset from inst/extdata
GSE77343 <- readRDS(system.file("extdata", "GSE77343.rds", package = "DeOPUS"))

# Inspect
dim(GSE77343$bulk)          # 17275 genes x 197 samples
dim(GSE77343$cellTypeExpr)  # 17275 genes x 15 cell types

# Run deconvolution (use multiple cores for speed)
results_gse <- deconvolve(
  bulk      = GSE77343$bulk,
  reference = GSE77343$cellTypeExpr,
  n_cores   = 4,
  verbose   = TRUE
)

# View proportions
head(results_gse$proportions)
```

### Compare With Available Ground Truth

```{r evaluate-real, eval=FALSE}
# GSE77343$bulkRatio contains measured proportions for 5 of the 15 cell types
common_types <- intersect(rownames(GSE77343$bulkRatio),
                          colnames(results_gse$proportions))

cor_per_type <- sapply(common_types, function(ct) {
  cor(results_gse$proportions[, ct],
      GSE77343$bulkRatio[ct, ],
      method = "pearson")
})

print(round(cor_per_type, 3))
```

## Method Details

DeOPUS applies a multi-level adaptive transformation:

### 1. Hierarchical Shrinkage

- **Local shrinkage** ($\tau$): Attenuates the influence of individual
  high-variance genes.
- **Global shrinkage** ($\phi$): Provides overall regularization across the
  expression profile.

### 2. Power Transformation

A variance-stabilizing power transformation handles the high dynamic range:

$$z = \frac{(q/s + \kappa)^{\mu \omega g} - \kappa^\mu}{\mu g}$$

where $q$ is the expression value, $s$ is a scale factor, $g$ is the global
shrinkage weight, and $\kappa$, $\mu$, $\omega$ are shape parameters.

### 3. Quantile Normalization

Spline-based quantile normalization ensures robust comparison between profiles.

### 4. Optimization

DeOPUS uses L-BFGS-B optimization with box constraints (proportions ≥ 0) to
minimize the loss between transformed bulk and reconstructed expression.

## Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `alpha`   | 0.01    | Regularization parameter |
| `power`   | 2       | Loss function power (1 = MAE, 2 = MSE) |
| `n_cores` | 1       | Parallel processing cores |
| `maxit`   | 100     | Maximum optimization iterations |
| `verbose` | FALSE   | Print progress messages |

## Session Info

```{r sessioninfo}
sessionInfo()
```
