---
title: "Getting Started with DeOPUS"
author: "Author Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DeOPUS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction
DeOPUS (Deconvolution via Optimized Power-transformed Unmixing with Shrinkage) is a reference-based cellular deconvolution method that estimates cell-type proportions from bulk RNA sequencing data.

## Installation

Install DeOPUS from GitHub:

```{r install, eval=FALSE}
devtools::install_github("yourusername/DeOPUS")
```

Load the package:

```{r load}
library(DeOPUS)
```

## Quick Example

### Prepare Input Data

DeOPUS requires two inputs:

1. **Reference matrix**: Average expression profiles for each cell type (genes × cell types)
2. **Bulk matrix**: Bulk RNA-seq expression data (genes × samples)

```{r data, eval=FALSE}
# Example: Load your data
# reference <- readRDS("path/to/reference.rds")
# bulk <- readRDS("path/to/bulk.rds")

# Create example data
set.seed(42)
n_genes <- 1000
n_cell_types <- 5
n_samples <- 10

# Simulated reference (genes x cell types)
reference <- matrix(
  rpois(n_genes * n_cell_types, lambda = 100),
  nrow = n_genes,
  ncol = n_cell_types
)
rownames(reference) <- paste0("Gene", 1:n_genes)
colnames(reference) <- paste0("CellType", 1:n_cell_types)

# True proportions
true_props <- matrix(
  runif(n_samples * n_cell_types),
  nrow = n_samples,
  ncol = n_cell_types
)
true_props <- true_props / rowSums(true_props)

# Simulated bulk (genes x samples) = reference %*% t(proportions) + noise
bulk <- reference %*% t(true_props) + matrix(rnorm(n_genes * n_samples, sd = 10), nrow = n_genes)
bulk[bulk < 0] <- 0
colnames(bulk) <- paste0("Sample", 1:n_samples)
```

### Run Deconvolution

```{r deconvolve, eval=FALSE}
# Run DeOPUS
results <- deconvolve(
  bulk = bulk,
  reference = reference,
  alpha = 0.01,
  n_cores = 4,
  verbose = TRUE
)

# View estimated proportions
print(results$proportions)
```

### Evaluate Results

```{r evaluate, eval=FALSE}
# Compare with true proportions
estimated <- results$proportions

# Calculate correlation
cor_values <- sapply(1:n_samples, function(i) {
  cor(estimated[i, ], true_props[i, ], method = "pearson")
})

cat("Mean Pearson correlation:", mean(cor_values), "\n")
```

## Method Details

DeOPUS applies a multi-level adaptive transformation:

### 1. Hierarchical Shrinkage

The transformation incorporates both local and global shrinkage:

- **Local shrinkage** ($\tau$): Attenuates the influence of individual high-variance genes
- **Global shrinkage** ($\phi$): Provides overall regularization across the expression profile

### 2. Power Transformation

A variance-stabilizing power transformation handles the high dynamic range:

$$z = \frac{(q/s + \kappa)^{\mu \omega g} - \kappa^\mu}{\mu g}$$

where $q$ is the expression value, $s$ is a scale factor, and $g$ is the global shrinkage weight.

### 3. Quantile Normalization

Spline-based quantile normalization ensures robust comparison between profiles:

```{r qnorm, eval=FALSE}
ranks <- rank(z) / length(z)
quantile_norm <- qnorm(pmax(0.001, pmin(0.999, ranks)))
```

### 4. Optimization

DeOPUS uses L-BFGS-B optimization with box constraints (proportions ≥ 0) to minimize the loss between transformed bulk and reconstructed expression.

## Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `alpha` | 0.01 | Regularization parameter |
| `power` | 2 | Loss function power (1=MAE, 2=MSE) |
| `n_cores` | 1 | Parallel processing cores |
| `maxit` | 100 | Maximum iterations |

## Benchmarking

To reproduce the benchmark results from our paper:

```{r benchmark, eval=FALSE}
# Run benchmark on real datasets
source("scripts/benchmark/run_benchmark_real.R")
run_benchmark()

# Generate visualization
source("scripts/analysis/visualize_results.R")
results <- aggregate_results()
dashboard <- create_summary_barplot_dashboard(results)
```

## Session Info

```{r sessioninfo}
sessionInfo()
```
